A Pandemic-Scale Ancestral Recombination Graph for SARS-CoV-2

### 2025-10-20 latex/main.tex
Draft of second paragraph

See Document~\ref{sec:detailed_ripples} for more detailed analysis 
of the UShER+RIPPLES results.
[

% TODO convert this into a discussion pargraph..
% \subsection*{Efficient inference and analysis of a pandemic-scale ARG}

% The COVID-19 pandemic resulted in the collection of viral genome sequence data on an unprecedented scale,
% which overwhelmed existing computational approaches,
% motivating the rapid development of new methods \cite{Turakhia2021,DeMaio2023,Hinrichs2024}.
% Although the computational tools used in phylogenetics and population genetics have historically been largely distinct,
% the functionality required largely overlaps.
% Sc2ts was built by modifying components of the tskit ARG software ecosystem \cite{Kelleher2019,Ralph2020,Baumdicker2022,Wong2024}
% and the VCF Zarr data storage format \cite{Czech2025}
% to describe relationships among virus samples,
% building an ARG over time as data becomes available.
% By reusing high-quality and high-performance components of these toolkits,
% we could develop the sc2ts inference method itself and
% perform all the pandemic-scale analyses in this article
% with Python code alone and no compiled languages.

% To illustrate the power of these libraries, we provide some examples
% in which we compare the Usher and sc2ts intersections in tskit format described earlier.
% First, we computed levels of concordance when imputing missing and ambiguous bases in the alignments.
% The total aligned dataset for 1,226,587 samples and 27,431 sites
% shared by the sc2ts and Usher tree sequences represents 31.34 GB of nucleotide calls.
% Of the 80,574,518 missing data calls (Ns) in the alignments,
% sc2ts and Usher disagreed in their imputed values for only 56,782 sites (0.07\%).
% Additionally, 413,303 calls made use of the IUPAC uncertainty codes.
% Of these sc2ts imputed only 77 (0.02\%) incorrectly
% (i.e., with a base that is not compatible with the ambiguity code).
% UShER imputed only 4 calls from this set incorrectly,
% reflecting its explicit incorporation of IUPAC uncertainty codes in parsimony calculations.
% This computation,
% performed by co-iterating over the sites in the alignments and the sc2ts and Usher trees,
% required 30 lines of Python code and took 2m 59s on an intel i7-9700 CPU,
% with a peak memory usage of about 4.5 GB
% (see \cite{Czech2025} for further benchmarks for
% processing the Viridian data in VCF Zarr format).
% Performing this analysis using matUtils would require
% exporting the full dataset to VCF, and
% converting the VCF to FASTA, and
% then comparing the alignments one-by-one to the source FASTA alignments.
% In contrast with matUtils and most phylogenetic software,
% which provide a command line interface and suite of predefined operations that typically output text files,
% tskit is a Python API (with C and Rust interfaces also supported)
% which has many advantages \cite{Baumdicker2022}.
% As an illustration of the open-ended and efficient nature of the analyses this enables,
% we computed the number of polytomies (nodes that have $> 2$ children)
% and the maximum number of children per node in the Usher and sc2ts outputs.
% In the Usher tree it took 116 ms and 4 lines of code to count polytomies (171,148)
% and compute the maximum number of children (5918).
% In the sc2ts ARG, it took 385 ms and 6 lines of code
% to find the maximum number of polytomies (137,001) and
% maximum number of children (7708) across all 348 local trees.

@@

We present here the first pandemic-scale Ancestral Recombination Graph (ARG)
for SARS-CoV-2, reconstructed from 2.48 million genomes. This resource provides
a comprehensive, recombination-aware view of the evolutionary history of the
virus, capturing not only point mutations but also deletions and recombination
events in a single coherent framework. By representing SARS-CoV-2 evolution as
a graph rather than a tree, the ARG unifies the relationships among lineages
that could not previously be reconciled by tree-based phylogenies. The
resulting structure serves as a long-term reference for studies of viral
evolution and as a foundation for future recombination-aware surveillance.

Our inference of recombination represents a conservative lower bound on the
true frequency of recombination in SARS-CoV-2. This conservatism arises from
stringent quality control criteria, including the requirement for multiple
supporting loci for each parental assignment. Even under these strict
thresholds, the ARG contains hundreds of high-confidence recombination events
with a validation rate exceeding 90\% among events passing QC filters. The
overall prevalence of recombination inferred here is substantially lower than
early estimates based on smaller or less curated datasets, but this is
consistent with our aim to prioritise specificity over sensitivity.
Importantly, at the default parsimony penalty ($p = 4$), the ARG recovers the
majority of previously recognised recombinant lineages, including those most
relevant to global transmission dynamics.

Comparisons with the Pango X designations reveal that several nominally
distinct recombinant lineages share common ancestral recombination events. This
over-splitting has led to an inflated perception of recombinant diversity.
Within Omicron, for example, multiple X lineages are better explained as
descendants of a single ancestral recombinant rather than as independent
crossover events. Resolving these redundancies clarifies long-standing
complexity in the Omicron genealogy and simplifies the interpretation of
SARS-CoV-2’s reticulate evolution. Conversely, several designated X lineages
show little evidence of recombination in the ARG, emphasising that not all
phylogenetically unusual lineages are true recombinants. Together, these
results provide a more parsimonious and internally consistent model of
recombinant evolution.

Using the dated ARG, we examined the temporal and epidemiological correlates of
recombination. The frequency of inferred recombination events correlates
strongly with both global case numbers and the diversity of circulating
lineages. This pattern supports the hypothesis that recombination is driven
jointly by viral load in the population, which increases opportunities for
co-infection, and by lineage diversity, which provides distinct parental
genomes. These findings imply predictable ``hot periods'' for recombinant
emergence during pandemics, particularly when multiple antigenically divergent
variants co-circulate at high prevalence. Understanding these dynamics may aid
future outbreak forecasting and genomic surveillance strategies.

Analysis of genomic breakpoint locations revealed enrichment of recombination
near the boundaries of the Spike gene, consistent with previous observations
from smaller datasets. The repeated occurrence of breakpoints in these regions
suggests that structural or functional constraints may influence where viable
recombinants can arise. Mechanistic biases in template switching, combined with
selective pressures on Spike for receptor binding and immune escape, may
together shape the landscape of recombination. The ARG thus not only captures
recombination events but also points to underlying biological mechanisms worthy
of experimental follow-up.

Beyond its biological insights, this study represents a methodological advance
in real-time, recombination-aware phylogenetic inference. The \texttt{sc2ts}
method integrates an efficient, simplified Li--Stephens model with incremental
sample addition, allowing scalable ARG construction from millions of genomes.
We validated the method against empirical data and literature-curated
recombinant lineages rather than simulated datasets, because the global
SARS-CoV-2 collection offers an unparalleled real-world benchmark.
Nevertheless, standardised simulations will remain valuable for refining models
and comparing future implementations. Our results demonstrate that the
\texttt{sc2ts} framework captures the essential genealogical structure of
SARS-CoV-2 while accommodating recombination in a computationally tractable
way.

The ARG produced here is embedded within the \texttt{tskit} and
\texttt{VCF-Zarr} ecosystems, enabling compact storage, rapid analysis, and
open sharing. This integration ensures that the resource is immediately usable
by the broader community for tasks such as mapping selective sweeps, tracing
recombinant origins, and evaluating lineage dynamics. The dataset and
accompanying workflows provide a transparent foundation for benchmarking future
methods and for extending ARG-based analyses to other pathogens.

In summary, this work demonstrates that recombination in SARS-CoV-2 can be
robustly inferred and represented at scale, transforming how we study the
evolution of rapidly diversifying viruses. The ARG provides a unified,
validated view of SARS-CoV-2 evolution that resolves the relationships among
recombinant lineages, links recombination frequency to epidemiological
dynamics, and highlights mechanistic constraints on recombination. More
broadly, it establishes that scalable ARG inference is now a practical and
powerful tool for pandemic genomics.

\subsection*{OUTLOOK}

The pandemic-scale ARG presented here establishes a new foundation for studying
viral evolution in real time. Beyond SARS-CoV-2, the \texttt{sc2ts} framework
can be directly adapted to other recombining pathogens such as influenza, HIV,
and rotaviruses, where complex genealogies have long limited evolutionary
analysis. Integrating the ARG with epidemiological and immunological data will
enable quantitative studies of how recombination interacts with selection,
immune escape, and transmission dynamics. As sequencing becomes increasingly
routine in global surveillance, efficient ARG inference will allow real-time
tracking of recombinant emergence and fitness, supporting faster risk
assessment and lineage designation. The combination of scalable inference, open
data standards, and community tools transforms recombination from a nuisance
factor into an analyzable dimension of viral evolution. We anticipate that this
resource will not only clarify the history of SARS-CoV-2 but also guide the
design of next-generation computational and experimental approaches to
understanding recombination-driven adaptation in RNA viruses.

>>>>

See Document~\ref{sec:detailed_ripples} for more detailed analysis 
of the UShER+RIPPLES results.

@@

% P1. The ARG. Unique resource. Emphasise utility of UShER and how 
% this has more.

We present the first pandemic-scale ancestral recombination graph (ARG) for
SARS-CoV-2, reconstructed from 2.48 million genomes using \texttt{sc2ts}. This
ARG represents a step change in how the evolutionary history of this virus can
be represented and analysed. Like the UShER phylogeny, it provides a
comprehensive and continuously updatable record of global genomic diversity,
but it extends this paradigm by embedding recombination directly into the
genealogy. The ARG unifies point mutations, deletions, and recombination events
within a single, compact data structure, enabling analyses that
were previously fragmented across multiple methods and data models. By encoding
reticulate ancestry explicitly rather than assuming a single tree, the ARG
provides a complete and internally consistent representation of SARS-CoV-2
evolution. In this way, the
ARG offers the same accessibility and scalability that made the UShER tree
indispensable during the pandemic, while expanding its scope to encompass the
full spectrum of evolutionary processes shaping SARS-CoV-2.

% P2. This is a conservative lower-bound on recombination. 
% Discuss other estimates.

The recombination frequency estimated from the \texttt{sc2ts} ARG is
substantially lower than that implied by the combination of UShER+RIPPLES,
largely reflecting differences in model formulation and parameter choice.
RIPPLES, applied post-hoc to UShER trees of 1.6 million genomes up to May 2021,
reported 589 unique recombination events with 43,104 descendant
samples---corresponding to roughly 2.7\% of all genomes being of recombinant
origin.  Using the same time interval, our inference at \(k=4\) identifies only
14 high-confidence recombination events with 172 descendants among 501,235
genomes (0.034\%), a 40- to 80-fold reduction.  This difference is largely
explained by the stricter parsimony penalty adopted here: increasing the
penalty from \(p=3\) (as used in the RIPPLES analysis) 
to \(k=4\) greatly reduces false
positives while retaining nearly all verified recombinant lineages, 
including those missed by RIPPLES at equivalent thresholds.  
We argue that the post-hoc nature of UShER+RIPPLES can erase evidence 
of recombination as
tree rearrangements proceed, thereby requiring the more lenient 
threshold in order to find known recombinats.
These comparisons suggest that our estimate represents a conservative 
lower bound. Further work will be required to determine
the true frequency of recombination in SARS-CoV-2 evolution.

% P3. Clarifies relationships between pango X lineages, simplifying
% the Omicron sublineage.

% P4. Results on global case numbers and frequency of recomb.

% P5. Confirm breakpoints around Spike. suggests further study.

% P6. Sc2ts method is a major advance. We validate with real data
% rather than simulations. Forward looking statement about simulations.

% P7. The software ecosystem enables diverse besplike analysis.

% P8. Outlook

\begin{itemize}
\item We present a pandemic-scale ancestral recombination graph (ARG) that
provides the first comprehensive, recombination-aware reconstruction of
SARS-CoV-2 evolutionary history. This ARG unifies mutational, deletional,
and recombinational processes within a single coherent framework and
serves as a long-term reference for viral evolution.

\item Our estimate of recombination frequency represents a conservative
lower bound, reflecting stringent quality control and a high validation
rate of detected events. Even at this conservative threshold, sc2ts
recovers the majority of previously recognized recombinant lineages,
indicating that recombination has been a recurrent yet tractable feature
of SARS-CoV-2 evolution.

\item By comparing to the Pango X designations, we show that several
recombinant lineages share common ancestral recombination events,
revealing systematic over-splitting of designations and clarifying the
relationships among Omicron sublineages. This synthesis simplifies the
previously complex picture of Omicron evolution and supports a model of
hierarchical recombination.

\item Using this cleaned and validated ARG, we show that the frequency of
recombination events is correlated with both global case numbers and viral
lineage diversity. These results suggest that recombination frequency is
driven jointly by viral load and co-circulating diversity, leading to
predictable ``hot periods'' during large outbreaks.

\item Breakpoints are enriched near the boundaries of the Spike gene,
consistent with previous reports and suggesting that structural and
functional constraints shape where viable recombinants arise. These
recurrent junctions point to mechanistic or selective pressures on Spike
that warrant further experimental study.

\item The sc2ts method represents a methodological advance that enables
real-time, recombination-aware genealogy inference at pandemic scale. We
validated its performance against empirical data and literature-curated
events rather than simulations, because the global SARS-CoV-2 dataset
provides a uniquely comprehensive real-world benchmark. Nonetheless,
standardized simulations will be valuable for future methodological
refinement.

\item The resulting ARG, built within the tskit and VCF-Zarr ecosystems,
is an open, efficient, and extensible resource that can underpin studies
of selection, recombination mechanisms, and global viral spread.
It provides a foundation for benchmarking future inference methods and
for integrative analyses of viral evolution.

\item Together, these results demonstrate that recombination in
SARS-CoV-2 can be robustly inferred and represented at scale, and that
such representations are critical for understanding the emergence and
spread of major variants. Scalable ARG inference has now become a
practical tool for pandemic genomics.
\end{itemize}

Our catalogue of recombination events is intentionally conservative. Stringent
quality-control filters (requiring multiple supporting loci per parent) produce
a high validation rate among QC-passing events (>\;90\%), and the set of
recombinants we report should be interpreted as a lower bound on true
recombination. At the default parsimony penalty (\(p=4\)) \texttt{sc2ts}
recovers the majority of well-characterised recombinant lineages, while
avoiding many false positives that arise in less-curated datasets. Differences
between our prevalence estimates and earlier reports largely reflect these
choices of data curation and thresholds, and underscore the importance of
rigorous QC when estimating recombination at pandemic scale.

Comparisons to Pango X designations clarify previously confusing features of
SARS-CoV-2 recombinant diversity. Several nominally distinct X lineages are
parsimoniously explained as descendants of the same ancestral recombination
event, indicating systematic over-splitting in some parts of the nomenclature.
Conversely, a number of designated X lineages show little evidence of
recombination in the ARG. Taken together, these observations simplify the
genealogy of Omicron and related groups: rather than many independent
crossovers, a smaller number of ancestral recombination events gave rise to
nested clusters of descendant lineages.

The dated ARG enables an explicit connection between epidemiology and
recombination. The weekly count of inferred recombination events tracks both
global case numbers and lineage diversity; the product of cases and expected
heterozygosity (nH) is the strongest single predictor of event counts. This
joint dependence is consistent with the mechanistic expectation that
recombination requires both frequent co-infections (high prevalence) and
genetically distinct co-infecting viruses (high diversity), and implies
predictable ‘‘hot periods’’ for recombinant emergence during major waves.

Genomically, breakpoints are enriched near the Spike boundaries, a pattern
observed previously at smaller scales. The repeated occurrence of junctions
around Spike suggests that mechanistic biases in template switching and
selective constraints on Spike jointly shape where viable recombinants arise.
These hotspot locations therefore provide concrete hypotheses for experimental
work aimed at understanding how recombination and selection interact to produce
successful hybrid genomes.

Methodologically, \texttt{sc2ts} demonstrates that scalable, real-time ARG
inference is tractable for pandemic-scale datasets. By combining a simplified
Li and Stephens HMM with incremental sample addition 
we detect and incorporate recombination in a timely manner. The alternative
of using tree-only updates with post-hoc detection of recombination leads
to both inaccurate trees and the loss of recombination signal over time.
We validated \texttt{sc2ts} against empirical benchmarks
and curated recombinant lineages; while standardised simulations remain useful
for method comparisons and stress-testing, the global SARS-CoV-2 collection
provides an unparalleled real-world testbed and motivated our emphasis on
empirical validation and stringent QC.

The ARG and the \texttt{sc2ts} workflows are made available within the
\texttt{tskit} and VCF-Zarr ecosystems, enabling efficient storage, rapid
querying, and wide reuse. This integration makes the resource immediately
actionable for tasks such as tracing recombinant origins, mapping selective
sweeps, and benchmarking future inference methods. As pathogen genomic
surveillance grows, ARG-based approaches can be incorporated into routine
pipelines to provide real-time, recombination-aware situational awareness.

More broadly, the present work shows that recombination — long treated as a
nuisance in large-scale pathogen genomics — can be robustly inferred and
represented at scale. The ARG turns recombination into an analyzable dimension
of viral evolution and lays a practical foundation for future studies that
integrate genomic, epidemiological and immunological data to quantify how
recombination interacts with selection, immunity and transmission. We
anticipate that \texttt{sc2ts} and the ARG will catalyse both methodological
development and biological discovery for SARS-CoV-2 and other recombining
pathogens.

% We present here the first pandemic-scale Ancestral Recombination Graph (ARG)
% for SARS-CoV-2, reconstructed from 2.48 million genomes. This resource provides
% a comprehensive, recombination-aware view of the evolutionary history of the
% virus, capturing not only point mutations but also deletions and recombination
% events in a single coherent framework. By representing SARS-CoV-2 evolution as
% a graph rather than a tree, the ARG unifies the relationships among lineages
% that could not previously be reconciled by tree-based phylogenies. The
% resulting structure serves as a long-term reference for studies of viral
% evolution and as a foundation for future recombination-aware surveillance.

% Our inference of recombination represents a conservative lower bound on the
% true frequency of recombination in SARS-CoV-2. This conservatism arises from
% stringent quality control criteria, including the requirement for multiple
% supporting loci for each parental assignment. Even under these strict
% thresholds, the ARG contains hundreds of high-confidence recombination events
% with a validation rate exceeding 90\% among events passing QC filters. The
% overall prevalence of recombination inferred here is substantially lower than
% early estimates based on smaller or less curated datasets, but this is
% consistent with our aim to prioritise specificity over sensitivity.
% Importantly, at the default parsimony penalty ($p = 4$), the ARG recovers the
% majority of previously recognised recombinant lineages, including those most
% relevant to global transmission dynamics.

% Comparisons with the Pango X designations reveal that several nominally
% distinct recombinant lineages share common ancestral recombination events. This
% over-splitting has led to an inflated perception of recombinant diversity.
% Within Omicron, for example, multiple X lineages are better explained as
% descendants of a single ancestral recombinant rather than as independent
% crossover events. Resolving these redundancies clarifies long-standing
% complexity in the Omicron genealogy and simplifies the interpretation of
% SARS-CoV-2’s reticulate evolution. Conversely, several designated X lineages
% show little evidence of recombination in the ARG, emphasising that not all
% phylogenetically unusual lineages are true recombinants. Together, these
% results provide a more parsimonious and internally consistent model of
% recombinant evolution.

% Using the dated ARG, we examined the temporal and epidemiological correlates of
% recombination. The frequency of inferred recombination events correlates
% strongly with both global case numbers and the diversity of circulating
% lineages. This pattern supports the hypothesis that recombination is driven
% jointly by viral load in the population, which increases opportunities for
% co-infection, and by lineage diversity, which provides distinct parental
% genomes. These findings imply predictable ``hot periods'' for recombinant
% emergence during pandemics, particularly when multiple antigenically divergent
% variants co-circulate at high prevalence. Understanding these dynamics may aid
% future outbreak forecasting and genomic surveillance strategies.

% Analysis of genomic breakpoint locations revealed enrichment of recombination
% near the boundaries of the Spike gene, consistent with previous observations
% from smaller datasets. The repeated occurrence of breakpoints in these regions
% suggests that structural or functional constraints may influence where viable
% recombinants can arise. Mechanistic biases in template switching, combined with
% selective pressures on Spike for receptor binding and immune escape, may
% together shape the landscape of recombination. The ARG thus not only captures
% recombination events but also points to underlying biological mechanisms worthy
% of experimental follow-up.

% Beyond its biological insights, this study represents a methodological advance
% in real-time, recombination-aware phylogenetic inference. The \texttt{sc2ts}
% method integrates an efficient, simplified Li--Stephens model with incremental
% sample addition, allowing scalable ARG construction from millions of genomes.
% We validated the method against empirical data and literature-curated
% recombinant lineages rather than simulated datasets, because the global
% SARS-CoV-2 collection offers an unparalleled real-world benchmark.
% Nevertheless, standardised simulations will remain valuable for refining models
% and comparing future implementations. Our results demonstrate that the
% \texttt{sc2ts} framework captures the essential genealogical structure of
% SARS-CoV-2 while accommodating recombination in a computationally tractable
% way.

% The ARG produced here is embedded within the \texttt{tskit} and
% \texttt{VCF-Zarr} ecosystems, enabling compact storage, rapid analysis, and
% open sharing. This integration ensures that the resource is immediately usable
% by the broader community for tasks such as mapping selective sweeps, tracing
% recombinant origins, and evaluating lineage dynamics. The dataset and
% accompanying workflows provide a transparent foundation for benchmarking future
% methods and for extending ARG-based analyses to other pathogens.

% In summary, this work demonstrates that recombination in SARS-CoV-2 can be
% robustly inferred and represented at scale, transforming how we study the
% evolution of rapidly diversifying viruses. The ARG provides a unified,
% validated view of SARS-CoV-2 evolution that resolves the relationships among
% recombinant lineages, links recombination frequency to epidemiological
% dynamics, and highlights mechanistic constraints on recombination. More
% broadly, it establishes that scalable ARG inference is now a practical and
% powerful tool for pandemic genomics.

% \subsection*{OUTLOOK}

% The pandemic-scale ARG presented here establishes a new foundation for studying
% viral evolution in real time. Beyond SARS-CoV-2, the \texttt{sc2ts} framework
% can be directly adapted to other recombining pathogens such as influenza, HIV,
% and rotaviruses, where complex genealogies have long limited evolutionary
% analysis. Integrating the ARG with epidemiological and immunological data will
% enable quantitative studies of how recombination interacts with selection,
% immune escape, and transmission dynamics. As sequencing becomes increasingly
% routine in global surveillance, efficient ARG inference will allow real-time
% tracking of recombinant emergence and fitness, supporting faster risk
% assessment and lineage designation. The combination of scalable inference, open
% data standards, and community tools transforms recombination from a nuisance
% factor into an analyzable dimension of viral evolution. We anticipate that this
% resource will not only clarify the history of SARS-CoV-2 but also guide the
% design of next-generation computational and experimental approaches to
% understanding recombination-driven adaptation in RNA viruses.
